\textbf{Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-08-07 submission/manuscript.Rmd
changes after meeting 2

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019, @flemer_oral_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex. This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @@wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used. Depending on the application of the model, researchers may choose to use different modeling approaches. For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.

@@

>>>>

---
title: '**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**'
csl: mbio.csl
fontsize: 11pt
output:
  word_document:
geometry: margin=1.0in
bibliography: references.bib
---

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a ML model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing analysis frameworks to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

Among microbiome researchers the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex. This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used. Depending on the application of the model, researchers may choose to use different modeling approaches. For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.

@@

Paragraph about deploying a model.

### 2019-08-07 submission/manuscript.tex
changes after meeting 2

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--11).
The task of diagnosing an individual with high confidence relies on a ML
model that is built with rigorous methods. However, ML methodology in
the microbiome field has flaws that have not been addressed previously.
There is a lack of transparency in which methods are used and how these
methods are implemented (12, 13); models are being developed and
evaluated without a seperate held-out test data (3, 4, 14--16); there is
large variation between the cross-validation outcomes (16, 17) and
between cross-validation and testing performances (18) . These practices
limit reproducibility, cause overoptimism for model performance and
prevent the development of generalizable models, where the model makes
accurate predictions with newly acquired data as well as it does with
the training data. Nevertheless, the microbiome field is making progress
to avoid potential pitfalls of ML. More and more studies are now
validating their models on independent datasets (7, 18, 19) and they are
introducing frameworks to accurately use ML tools (20--23).

Among microbiome researchers, there has been a trend towards using more
complex ML models such as random forest and neural networks (2, 11,
24--26) over simpler models such as logistic regression or other linear
models (18, 22, 27). There is an implicit assumption that the more
complex models are better because they are more complex. Although these
models may be better at incorporating non-linear relationships or yield
better predictions, they are also called black box ML models because
they are not inherently interpretable. These models require post hoc
explanations to quantify the importance of each feature in making a
prediction. Depending on the application of the model researchers may
chose to use different modeling approaches. For example, a researcher
trying to identify the populations causing a disease would likely want a
more interpretable model whereas a clinician may emphasize performance.
Although one may feel that they are sacrificing interpretability for
performance, that tradeoff may be minimal (28, 29).

The lack of transparency in model selection and using flawed modelling
and interpretation methods negatively impacts model validity. To
showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches using the same dataset and
pipeline. We established modeling pipelines for three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also developed four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest and XGBoost. We compared
the predictive performance, interpretability and computational
efficiency of the seven models to highlight the importance of model
selection. To demonstrate the performance of these modelling approaches
and our pipeline, we used data from a previously published study that
sought to classifiy individuals as having normal colons or colonic
lesions based on the 16S rRNA gene sequences collected from fecal
samples (3). This dataset was selected because it is a relatively large
collection of samples (N=490) connected to clinically significant
disease where there is ample evidence that it is driven by variation in
the microbiome (1, 3, 4, 30). With this dataset, we established
standards for ML pipeline construction, evaluated predictive
performance, and demonstrated model interpretation for models. This
framework can be easily applied to other host-associated and
environmental microbiome problems.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we train and validate each of the seven models {[}Figure
1{]}. We randomly split the data into training and test sets so that the
training set consisted of 80\% of the full dataset while the test set
was composed of the remaining data {[}Figure 1{]}. Since the cases are
not uniformly represented in the data, the data-split was stratified to
maintain the overall label distribution in both the training and test
sets. For example, our example dataset represented data from 490
individuals, 261 had normal colons and 229 had a screen relevant
neoplasia (SRN). After the data-split, the training set consisted of 393
patients (209 SRN), while the test set was composed of 97 patients (52
SRN). The training data was used for hyperparameter selection and the
test set was used for evaluating predictive performance.

We used ML models with different classification algorithms and
regularization methods. Regularization is a technique to discourage
overfitting by penalizing the model for learning the training data too
well. For regularized logistic regression and SVM with linear kernel, we
used L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero). Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGBoost, as well as an SVM with non-linear kernel.

Hyperparameters are the rules that are learned from the training data in
a classification algorithm. For example regularization term (C) that
decides how big the penalty for overfitting will be, is an
hyperparameter. The model becomes more generalizable when we tune for
the optimal C value. Similar to regularization term C, other
hyperparameters can be tuned over a full grid search and selected by
validation to build better models. We selected hyperparameter settings
by performing repeated five-fold cross-validation (CV) on the training
set {[}Figure 1{]}. Similar to the initial data-split, five-fold CV was
also stratified to maintain the overall label distribution. We chose the
best hyperparameter setting for each model based on its CV predictive
performance using the area under the receiver operating characteristic
curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC ranges from 1.0,
where the model perfectly distinguishes between cases and controls, to
0.50, where the model's predictions are no different from random chance.
The cross-validation of each hyperparameter setting was repeated over
100 randomizations to get a robust reading of predictive performance.

We then trained the full training dataset with the selected
hyperparameters. We used the held-out test set to evaluate the testing
predictive performance of each ML model. The data-split, hyperparameter
selection, training and testing steps were repeated 100 times to get a
reliable and robust reading of model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performances of seven binary
classification models when applied to held-out test data using the AUROC
metric {[}Figure 2{]}. Random forest had significantly higher test AUROC
values than the other models for detecting SRNs when AUROC values were
compared to the other six by Wilcoxon rank sum test (p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Random forest had the highest median AUROC for detecting
SRN. Despite its simplicity, the L2-regularized logistic regression was
second best in predictive performance.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. The difference
between the two should be low to suggest the model is not overfitting
despite the large number of features. The largest difference between the
two was 0.021 in L1-regularized SVM with linear kernel, followed by SVM
with radial basis function kernel and decision tree with a difference of
0.007 and 0.006, respectively {[}Figure 2{]}. We also reported the
testing AUROC values over 100 randomizations of the initial data-split.
The testing AUROC values within each model varied 0.23 on average across
the seven models. For instance, the lowest AUROC value of the random
forest model was 0.59 whereas the highest was 0.81. These results showed
that depending on the data-split, the testing AUROC values showed great
variability {[}Figure 2{]}.

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. The ML models we
built using L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels, a decision
tree, random forest and XGBoost decrease in interpretability as they
increase in complexity. In this study we highlighted two methods to
interpret models with varying complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear
kernel and L2-regularized logistic regression) using the absolute
feature weights of the trained models. We ranked the absolute weights of
all the OTUs for each data-split {[}Figure 3{]}. We calculated the
median ranks of these features over the 100 data-splits. In the three
linear models, OTUs that had the largest median ranks and drove the
detection of SRNs belonged to families \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
3{]}. Some of the OTUs with the highest ranks were shared among the
linear models.

>>>>

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a ML model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing analysis frameworks to better use ML tools (20--23). More
work is needed to further improve reproducibility and minimize
over-optimism for model performance.

Among microbiome researchers the lack of transparency in justifying a
modelling approach has been due to an implicit assumption that more
complex models are better because they are more complex. This has
resulted in a trend towards using models such as random forest and
neural networks (2, 11, 24--26) over simpler models such as logistic
regression or other linear models (18, 22, 27). Although the more
complex models may be better at incorporating non-linear relationships
or yield better predictions, they are considered to be black box models
because they are not inherently interpretable. These models require post
hoc explanations to quantify the importance of each feature in making a
prediction and they do not show the structure of how the features are
used. Depending on the application of the model, researchers may choose
to use different modeling approaches. For example, researchers trying to
identify the populations causing a disease would likely want a more
interpretable model whereas clinicians may emphasize performance.
Although one may feel that they are sacrificing interpretability for
performance, that tradeoff may be minimal (28, 29). Regardless, it is
important for researchers to articulate why they have selected a
specific modelling approach or even compare multiple approaches in the
same study.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also built four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared the
predictive performance, interpretability, and computational efficiency.
To demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a framework
that implementf a ML pipeline that can be used for any modeling
approach, evaluates predictive performance, and demonstrates how to
interpret these models. This framework can be easily applied to other
host-associated and environmental microbiome datasets.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we trained and validated each of the seven models using a
common approach that is based on standard methods within the ML
community (REFS){[}Figure 1{]}.

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset while the test
set was composed of the remaining 20\% of the data {[}Figure 1{]}. To
maintain the distribution of controls and cases that was found with the
full dataset, we performed stratified splits. For example, our full
dataset included 490 individuals. Of these, 261 had normal colons (53\%)
and 229 had a screen relevant neoplasia (SRN; 46.7\%). A training set
included 393 individuals, of which 209 had an SRN (53\%), while the test
set was composed of 97 individuals of which 52 had an SRN (54\%). The
training data was used to build the models and the test set was used for
evaluating predictive performance.

Second, we trained seven different models using the training data. We
selected models with different classification algorithms and
regularization methods. Regularization is a technique that discourages
overfitting by penalizing the model for learning the training data too
well. For regularized logistic regression and SVM with linear kernel, we
used L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data led to a sparser solution
(i.e., force many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including decision tree, random forest, and
XGBoost and we trained an SVM model with non-linear kernel.

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance.

Finally, we trained the full training dataset with the selected
hyperparameter values and applied the model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to get a reliable and robust reading of model
performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularlized logistic regression and
linear SVM) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs is large, then that
would indicate that the models were overfit. difference in median AUROCs
was 0.021 in L1-regularized SVM with linear kernel, followed by SVM with
radial basis function kernel and decision tree with a difference of
0.007 and 0.006, respectively {[}Figure 2{]}; however, these differences
are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range for each model using 100 splits. The range among the testing AUROC
values within each model varied by 0.23 on average across the seven
models. If we had only done a single split, then there is a risk that we
could gotten lucky or unlucky with the performance of the model. For
instance, the lowest AUROC value of the random forest model was 0.59
whereas the highest was 0.81. These results showed that depending on the
data-split, the testing AUROC values showed great variability {[}Figure
2{]}. Therefore, it is important to employ the hierarchical data splits
that were included in our pipeline to minimize the risk of
over-optimism.

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to interpret models with varying complexity.

We interpreted linear models (L1 and L2-regularized SVM with linear
kernel and L2-regularized logistic regression) using the absolute
feature weights of the trained models. We ranked the absolute values of
the weights of all the OTUs for each data-split {[}Figure 3{]}. We
calculated the median ranks of these features over the 100 data-splits.
In the three linear models, the OTUs that had the largest median rank
and drove the detection of SRNs contained sequences that were most
similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia
(OTU00822, OTU00768), Clostridium (OTU00609), Blautia (OTU00659,
OTU00012), Faecalibacterium (OTU00015), Gamella (OTU00426) and
Peptostreptococcus (OTU00367) {[}Figure 3{]}. Among the five highest
ranked OTUs across these three models, it was encouraging that OTU 1239
was shared among each of the models and that \ldots{} were shared among
both of the L2-regularized models.
